Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-03-20 overview.tex
Many minor changes.  No more undefined references!  Added appendix file.
more changes to intro, overview

\subsubsection{Magnetohydrodynamics and Gravity}

@@

\begin{eqnarray} 
\frac{\partial \rho}{\partial t} 
          + \frac{1}{a} \nabla \cdot (\rho \vecv) & = & 0
        \label{eq:mass} \\
%
\frac{\partial \rho \vecv}{\partial t}  
   + \frac{1}{a} \nabla \cdot \left(\rho \vecv \vecv + \myvec{I}p^* - \frac{\vecB \vecB}{a} \right) & = &
    - \frac{\dot{a}}{a} \rho \vecv - \frac{1}{a} \rho \nabla \phi
    %
%          + \frac{1}{a} ( \vecv \cdot \nabla ) \vecv
%    & = & - \frac{\dot{a}}{a} \vecv 
%          - \frac{1}{a \rho} \nabla p
%          - \frac{1}{a \rho} (\nalba \times \vecB) \times \vecB
%          - \frac{1}{a} \nabla \phi, 
        \label{eq:momentum} \\
%
\frac{\partial E}{\partial t} 
+ \frac{1}{a} \nabla \cdot \left[ (E + p^*) \vecv -  \frac{1}{a} \vecB (\vecB \cdot \vecv) \right] & = &
     - \frac{\dot{a}}{a} \left( 2E - \frac{B^2}{2a} \right)  - \frac{1}{a} \vecv \cdot \nabla \phi 
     %
%         + \frac{1}{a} \vecv \cdot \nabla E
 %  & = & - \frac{\dot{a}}{a} (3 \frac{p}{\rho} + {\vecv \cdot \vecv} )
  %       - \frac{1}{a \rho } \nabla \cdot (p \vecv) 
%%\nonumber   & &   %uncomment for 2-column
 %        - \frac{1}{a} \vecv \cdot \nabla \phi 
 %               + \Gamma - \Lambda.
       \label{eq:total_energy}  \\
%
\frac{\partial \vecB}{\partial t} - \frac{1}{a}  \nabla \times (\vecv \times \vecB) & = & 0 \label{eq:induction}
\end{eqnarray}

%\begin{eqnarray} 
%\frac{\partial \rho}{\partial t} 
%          + \vecv \cdot \nabla \rho
%    & = & - \rho \nabla \cdot \vecv,
%        \label{eq:mass} \\
%%
%\frac{\partial \vecv }{ \partial t} 
%          + ( \vecv \cdot \nabla ) \vecv 
%    & = & - \frac{1}{\rho} \nabla p
%          - \nabla \phi,
%        \label{eq:momentum} \\
%%
% \frac{\partial E}{\partial t} + \vecv \cdot \nabla E
%    & = & - \frac{1}{\rho} \nabla \cdot (p \vecv) 
%          - \vecv \cdot \nabla \phi
%%\nonumber          \\  & &     % uncomment for 2-column
%     + \Gamma - \Lambda
%        \label{eq:total_energy}
%\end{eqnarray}
%
Here $E$, $\rho$, $\vecv$, and $\vecB$ are the comoving total fluid energy, comoving baryonic density, peculiar velocity, and comoving magnetic field strength, respectively.  $\myvec{I}$ is the identity matrix.
The first equation represents conservation of mass, the second, conservation of momentum, and the third, conservation of total fluid (kinetic plus thermal) energy.  They are respectively, the first, second and third moments of the Boltzmann equation.  The fourth equation is the magnetic induction equation.   Terms representing radiative cooling ($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the energy equation (\ref{eq:total_energy}).

The total fluid energy $E$ is given by:
\begin{equation}
E =  \rho e + \frac{\rho v^2}{2}  + \frac{B^2}{2a}
        \label{eq:total_energy_def}.
\end{equation}
where $e$ is the comoving specific thermal energy, and the total isotropic pressure $p^*$ is given by
\begin{equation}
p^* = p + \frac{B^2}{2a}
\end{equation}
The equations are closed with the equation of state and Poisson's equation (we work in the Newtonian limit):
%
\begin{eqnarray}
%
e   & = & p / [(\gamma - 1) \rho ],
        \label{eq:eq_of_state} \\
%
\nabla^2 \phi & = & 4 \pi G \rho_{total}
%                   + 3 p_{total}/c^2) - \Lambda. 
        \label{eq:potential}
\end{eqnarray}
%
The equation of state is shown here for an ideal gas with a ratio of specific heats $\gamma$.
Equation (\ref{eq:potential}) includes contributions to the gravitational potential $\phi$ from the total mass density $\rho_{\rm total}$.  Although we write the equations including both MHD and comoving coordinates, the code can operate both in the pure hydrodynamic limit, and without cosmological expansion.

@@

\subsubsection{Comoving Coordinates}

@@

\subsubsection{Particles and Chemistry}

@@

In addition, we solve mass conservation equations for any chemical species.  For any species $i$, these equations have the form:

\begin{eqnarray}
\frac{\partial \rho_i}{\partial t} 
          + \frac{1}{a} \vecvb \cdot \nabla \rho_i 
     & = &  - \frac{1}{a} \rho_i \nabla \cdot \vecvb 
        + \sum_j \sum_l k_{jl}(T) \rho_j \rho_l 
% \nonumber \\    & &    %uncomment for 2-column
      + \sum_j I_j \rho_j 
        \label{eq:mass_species}
\end{eqnarray}
where $k_{ij}$ are the rate coefficients for the two-body reactions and are usually functions of only the gas species (we will specifically note the cases where we either include three-body reactions or have density-dependent rates).  The $I_j$ are destruction/creation rates due to photoionizations and/or photodissociations.

@@

\subsubsection{Radiation transport: ray-tracing}

\subsubsection{Radiation transport: moment method}

\subsubsection{Thermal conduction}

@@

\begin{equation}
\mathbb{F} = -\kappa_{sp} \nabla T
\end{equation}

@@

\begin{equation}
\mathbb{F} = -\kappa_{sp} \hat{b} (\hat{b} \cdot \nabla T)
\end{equation}

@@

In this section, we briefly describe the numerical methods which are used to solve the equations outline above.  We sweep through in the same order as will be used in section~\ref{sec.methods} so that there is a one-to-one correspondence between each of the following overviews and the complete description provided in section~\ref{sec.methods}.  The goal here is to introduce the reader to the basic idea of the methods without drowning in detail.

@@

\subsubsection{Godunov PPM method (Hydrodynamics)}

@@

PPM is an explicit, higher order-accurate version of
Godunov's method for ideal gas dynamics with third order-accurate piecewise parabolic
monotonic interpolation and a nonlinear Riemann solver for shock
capturing.  It advances the hydrodynamic equations in the following
steps:
\begin{enumerate}
 \item Construct monotonic parabolic interpolation of cell average
 data, for each fluid quantity.
 \item Compute interface states by averaging the parabola for over the domain of dependence for
 each interface
 \item Use interface data to solve the Riemann problem.
 \item Difference the interface fluxes to update the cell average quantities.
\end{enumerate}
It does an excellent job of capturing strong shocks in at
most two cells.  Multidimensional schemes are built up by directional
splitting and produce a method that is formally second order-accurate
in space and time which explicitly conserves mass, linear momentum, and energy.

@@

\subsubsection{Godonov MUSCL with Dedner (HD and MHD)}

\subsubsection{MHDCT}

\subsubsection{Second order finite difference method (HD)}

@@

\subsubsection{Chemistry, radiation and cooling}

@@

>>>>

%\subsubsection{Magnetohydrodynamics and Gravity}

@@

\begin{eqnarray} 
\frac{\partial \rho}{\partial t} 
          + \frac{1}{a} \nabla \cdot (\rho \vecv) & = & 0
        \label{eq:mass} \\
%
\frac{\partial \rho \vecv}{\partial t}  
   + \frac{1}{a} \nabla \cdot \left(\rho \vecv \vecv + \myvec{I}p^* - \frac{\vecB \vecB}{a} \right) & = &
    - \frac{\dot{a}}{a} \rho \vecv - \frac{1}{a} \rho \nabla \phi
    %
%          + \frac{1}{a} ( \vecv \cdot \nabla ) \vecv
%    & = & - \frac{\dot{a}}{a} \vecv 
%          - \frac{1}{a \rho} \nabla p
%          - \frac{1}{a \rho} (\nalba \times \vecB) \times \vecB
%          - \frac{1}{a} \nabla \phi, 
        \label{eq:momentum} \\
%
\frac{\partial E}{\partial t} 
+ \frac{1}{a} \nabla \cdot \left[ (E + p^*) \vecv -  \frac{1}{a} \vecB (\vecB \cdot \vecv) \right] & = &
     - \frac{\dot{a}}{a} \left( 2E - \frac{B^2}{2a} \right)  - \frac{1}{a} \vecv \cdot \nabla \phi 
     - \Lambda + \Gamma + \frac{1}{a^2} \nabla \cdot \fcond
     %
%         + \frac{1}{a} \vecv \cdot \nabla E
 %  & = & - \frac{\dot{a}}{a} (3 \frac{p}{\rho} + {\vecv \cdot \vecv} )
  %       - \frac{1}{a \rho } \nabla \cdot (p \vecv) 
%%\nonumber   & &   %uncomment for 2-column
 %        - \frac{1}{a} \vecv \cdot \nabla \phi 
 %               + \Gamma - \Lambda.
       \label{eq:total_energy}  \\
%
\frac{\partial \vecB}{\partial t} - \frac{1}{a}  \nabla \times (\vecv \times \vecB) & = & 0 \label{eq:induction}
\end{eqnarray}

%\begin{eqnarray} 
%\frac{\partial \rho}{\partial t} 
%          + \vecv \cdot \nabla \rho
%    & = & - \rho \nabla \cdot \vecv,
%        \label{eq:mass} \\
%%
%\frac{\partial \vecv }{ \partial t} 
%          + ( \vecv \cdot \nabla ) \vecv 
%    & = & - \frac{1}{\rho} \nabla p
%          - \nabla \phi,
%        \label{eq:momentum} \\
%%
% \frac{\partial E}{\partial t} + \vecv \cdot \nabla E
%    & = & - \frac{1}{\rho} \nabla \cdot (p \vecv) 
%          - \vecv \cdot \nabla \phi
%%\nonumber          \\  & &     % uncomment for 2-column
%     + \Gamma - \Lambda
%        \label{eq:total_energy}
%\end{eqnarray}
%
Here $E$, $\rho$, $\vecv$, and $\vecB$ are the comoving total fluid energy, comoving baryonic density, peculiar velocity, and comoving magnetic field strength, respectively.  $\myvec{I}$ is the identity matrix.
The first equation represents conservation of mass, the second, conservation of momentum, and the third, conservation of total fluid (kinetic plus thermal) energy.  They are respectively, the first, second and third moments of the Boltzmann equation.  The fourth equation is the magnetic induction equation.   Terms representing radiative cooling ($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the energy equation (\ref{eq:total_energy}), as well as the flux due to thermal heat conduction ($\fcond$).

The total fluid energy $E$ is given by:
\begin{equation}
E =  e + \frac{\rho v^2}{2}  + \frac{B^2}{2a}
        \label{eq:total_energy_def}.
\end{equation}
where $e$ is the comoving thermal energy density, and the total isotropic pressure $p^*$ is given by
\begin{equation}
p^* = p + \frac{B^2}{2a}
\end{equation}
The equations are closed with the equation of state and Poisson's equation (we work in the Newtonian limit):
%
\begin{eqnarray}
%
e   & = & p / [(\gamma - 1) \rho ],
        \label{eq:eq_of_state} \\
%
\nabla^2 \phi & = & 4 \pi G \rho_{\rm total}
%                   + 3 p_{total}/c^2) - \Lambda. 
        \label{eq:potential}
\end{eqnarray}
%
The equation of state is shown here for an ideal gas with a ratio of specific heats $\gamma$.
Equation (\ref{eq:potential}) includes contributions to the gravitational potential $\phi$ from the total mass density $\rho_{\rm total}$.  Although we write the equations including both MHD and comoving coordinates, the code can operate both in the pure hydrodynamic limit, and without cosmological expansion.

@@

%\subsubsection{Comoving Coordinates}

@@

%\subsubsection{Particles}

@@

%\subsubsection{Chemistry}

In addition, we solve mass conservation equations for any chemical species, including a variety of chemical networks.  For any species $i$ with comoving number density $n_i$, these equations have the form:
\begin{equation}
\frac{\partial n_i}{\partial t} 
          + \frac{1}{a} \nabla \cdot (n_i \vecv) = 
        k_{ij}(T) n_i n_j 
      + \Gamma^{ph}_j n_j 
        \label{eq:species_evolution}
\end{equation}
where $k_{ij}$ are the rate coefficients for the two-body reactions and are usually functions of only the gas species (we will specifically note the cases where we either include three-body reactions or have density-dependent rates).  The $\Gamma^{ph}_j$ are destruction/creation rates due to photoionizations and/or photodissociations.

@@

%\subsubsection{Radiation transport: ray-tracing}

The code can include either a homogeneous radiation background (with possible local self-shielding corrections), or evolve an inhomogeneous radiation field either by directly solving the radiative transfer equation along rays, or by solving a set of moment equations derived from the radiative transfer equation.

The radiative transfer equation in comoving coordinates 
\citep[e.g.,][]{Gnedin97} reads
%
\begin{equation}
  \label{eq:rteqn}
  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
  \frac{\hat{n} \cdot \nabla I_\nu}{\bar{a}} -
  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is the radiation
specific intensity in units of energy per time $t$ per solid angle per
unit area per frequency $\nu$.  $H = \dot{a}/a$ is the Hubble
constant, where $a$ is the scale factor.  $\bar{a} = a/a_{em}$ is the
ratio of scale factors at the current time and time of emission.  The
second term represents the propagation of radiation, where the factor
$1/\bar{a}$ accounts for cosmic expansion.  The third term describes
both the cosmological redshift and dilution of radiation.  On the
right hand side, the first term considers the absorption coefficient
$\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$, and the second term
$j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ is the emission coefficient
that includes any point sources of radiation or diffuse radiation.

%\subsubsection{Radiation transport: moment method}

In the other limit, we solve an equation derived from integrating
the equation of radiative transfer over angles.  In particular, we use a
flux-limited diffusion approximation for cosmological radiative
transfer, with couplings to both the gas energy and chemical number
densities. The equations solved are:
\begin{eqnarray}
  \label{eq:fld_radiation}
  \frac{\partial E_r}{\partial t} + \frac1a \nabla\cdot\left(E_r \vecv \right) &=& 
    \nabla\cdot\left(D\nabla E_r\right) -
    \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
  \label{eq:fld_heating}
  \frac{\partial e_c}{\partial t} &=& -\frac{2\dot{a}}{a} e_c + \Gamma - \Lambda,
%  \label{eq:fld_chemistry}
%  \partial_t {\mathrm n}_i + 
%    \frac1a \nabla\cdot\left({\mathrm n}_i \vec{v}_b\right) &=& 
%    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
%    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
\end{eqnarray}
where $E_r$ is a grey radiation energy density, $e_c$ is the
internal energy correction due to photo-heating and chemical cooling.
%and ${\mathrm n}_i$ correspond to the chemical number densities of
%HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
%number density.  
Here, we define the grey radiation energy density
through first assuming a fixed frequency spectrum, i.e.
$E_{\nu}(\nu,\vec{x},t) = \tilde{E}_r(\vec{x},t) \chi(\nu)$, and then
defining the integrated quantity
\begin{equation}
\label{eq:grey_radiation_energy}
   E_r(\vec{x},t) \equiv \int_{\nu_0}^{\infty}
   E_{\nu}(\nu,\vec{x},t)\,\mathrm d\nu \  = \ 
   \tilde{E}_r(\vec{x},t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
\end{equation}
In addition, $D$ in (\ref{eq:fld_radiation}) is the {\em Larsen
square-root flux-limiter} (see \cite{Morel2000}), $\kappa$ is the total
opacity, $\eta$ is the field of radiation sources, $\Gamma$ provides the
radiation induced photo-heating and, $\Lambda$ the chemical
cooling.
%, $\alpha_{i,j}$ are the reaction rate coefficients defining
%the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
%and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

%\subsubsection{Thermal conduction}

@@

\begin{equation}
\fcond = -\kappa_{sp} \nabla T
\end{equation}

@@

\begin{equation}
\fcond= -\kappa_{sp} \hat{b} (\hat{b} \cdot \nabla T)
\end{equation}

@@

In this section, we briefly describe the numerical methods which are used to solve the equations outlined above.  We sweep through in the same order as will be used in section~\ref{sec.methods} so that there is a one-to-one correspondence between each of the following overviews and the complete description provided in section~\ref{sec.methods}.  The goal here is to introduce the reader to the basic idea of the methods without drowning in detail.

@@

\subsubsection{Godunov PPM method (HD only)}

@@

PPM is an explicit, higher order-accurate version of
Godunov's method for ideal gas dynamics with third order-accurate piecewise parabolic
monotonic interpolation and a nonlinear Riemann solver for shock
capturing.  It advances the hydrodynamic equations in the following
steps: 
%\begin{enumerate}
% \item 
 (i) Construct monotonic parabolic interpolation of cell average data, for each fluid quantity.
% \item 
(ii) Compute interface states by averaging the parabola for over the domain of dependence for each interface
% \item 
(iii) Use interface data to solve the Riemann problem.
 %\item 
 (iv) Difference the interface fluxes to update the cell average quantities.
%\end{enumerate}

It does an excellent job of capturing strong shocks in at
most two cells.  Multidimensional schemes are built up by directional
splitting and produce a method that is formally second order-accurate
in space and time which explicitly conserves mass, linear momentum, and energy.

@@

\subsubsection{Godunov MUSCL (HD) with Dender divergence cleaning (MHD)}

\subsubsection{MHD with Constrained Transport (MHD)}

\subsubsection{Second order finite difference method (HD only)}

@@

\subsubsection{Chemistry}
\label{sec.ov.chem}

\subsubsection{Radiative Cooling and Heating}

@@

\subsubsection{Homogeneous radiation backgrounds}

\subsubsection{Radiation transport: ray tracing}

\subsubsection{Radiation transport: Flux-limited diffusion}

\subsubsection{Thermal Heat Conduction}
